Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2017-07-19 cabios-template/main.tex
in preprocessing of results

\abstract{\textbf{Motivation:} Genome-wide genetic studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge for the genetic community, severely slowing down genetic analyses. This increase in data size has led to the emergence of a range of specialized software for every part of the analysis pipeline. Yet, it is often difficult to choose which piece of software to use and how to combine all these software. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of the RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the Celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of our packages by analyzing a simulated Genome-Wide dataset including 500,000 individuals and 1M markers on a single desktop computer. \\
\textbf{Availability:} https://privefl.github.io/bigstatsr/ and https://privefl.github.io/bigsnpr/\\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools available either independently or incorporated within large framework (VCFtools, BCFtools, PLINK...). Similarly genome-wide analysis quality control software have been developed (e.g. Bioconductor GWASTools and PLINK). Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA (4 REFs). Then, GWAS analysis itself depends on the genotype format (ProbABEL or SNPTEST for dosage data).  Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred, PRSice and other. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

@@

The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory. It uses what we could call an hybrid solution: memory-mapping. Memory-mapping is used to access data, possibly stored on disk, as if it were in memory. This solution is made available within R through an object called ``big.matrix'', available in R package bigmemory (\cite{Kane2013}).

We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ (REF). With the two packages we developed, we made this solution available in R and in C++ via package Rcpp (REF). The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra (REFs).

Usually, memory-mapping provides faster and seamless access than standard read/write operations. When some element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. All this is managed by the Operating System so that it is seamless and efficient. It means that if you use chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on. Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in the memory and every second access would be fast.

@@

Fast read/write operations from/to bed/bim/fam PLINK files are available. One should use PLINK to convert data to this format. In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure 1). PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.

@@

Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure D). The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This algorithm doesn't use phasing and is very fast. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions (REF). XGBoost build decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices. For each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets. The training set is used to build a model for predicting missing data. The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of individuals that have been wrongly imputed for that particular SNP.

@@

The function based on randomized projections runs in linear time in all dimensions (REF). FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra (REF). The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes the multiplication based on the ``big.matrix'' format, which enables parallel computations.

Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed. For instance, thinning is mandatory when computing PCs of a genotype matrix (REF). There are different options to thin SNPs based on Linkage Disequilibrium. The first option is known as pruning, which is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning.

A variant of pruning is clumping. Clumping is useful if a statistic is available to sort the SNPs by importance (e.g. association with a phenotype) and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest. Furthermore, we advise to always use clumping instead of pruning (by using the MAF as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all (URL).

The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions (REF). Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in these regions while computing PCs (URL). However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and remove them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt (REF). Finally, the algorithm keeps only consecutive outlier SNPs which is considered as the sign of long-range LD regions by the algorithm. Indeed, a long-range LD region would cause SNPs in this region to have a strong weight (loadings) in the PCA and we can differentiate these from true outliers because they are consecutive. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

For association purposes, statistical tests based on linear and logistic regressions are available. Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome (REF).

$\forall j \in \{1, \dots, m\}$, 
\itemize{
\item for the linear regression: $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $m$ is the number of SNPs, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as Age or Gender).
\item for the logistic regression: $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K$, where $\hat{p} = \mathbb{P}(Y = 1)$.
}\\
The hypothesis that is tested is $\beta^{(j)} = 0$ against the alternative $\beta^{(j)} \neq 0$.

@@

First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in equations X and Y (REF evans2009). Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients.

Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors. The currently available models are linear and logistic regressions and Support Vector Machine (SVM). These models include lasso (REF) and elastic-net (REF) regularizations, which reduce the number of  predictors (SNPs) included in the predictive models. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors (REF). The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso (REF). These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).

@@

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compared the performance of the packages against standard software. Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: Celiac, a real case-control cohort, POPRES, a general population cohort and simulated datasets using real genotypes from the Celiac cohort. We compare the performance on two computers, a desktop computer with 64GB of memory and 12 cores, and a laptop with only 8GB of memory and 4 cores. For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort. The data have been pre-processed following steps from figure 1, resulting in 15,250 individuals and 281,122 SNPs. We note that if we used a standard R matrix to store the genotypes, it would take 32GB of memory. On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB.

@@

\subsection{Imputation of missing values for genotyped SNPs}

@@

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Tue Jun 13 11:25:56 2017
\begin{table}[ht]
\centering
\begin{tabular}{l|cccccccccc}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 \\
  \hline
PC1 & 100.0 & -0.1 & -0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
  PC2 & 0.1 & 100.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 \\
  PC3 & 0.1 & -0.0 & 99.9 & 0.2 & -0.0 & 0.1 & 0.1 & 0.1 & 0.0 & -0.1 \\
  PC4 & -0.0 & -0.0 & -0.3 & 99.9 & -0.1 & 0.1 & -0.1 & 0.0 & 0.1 & 0.1 \\
  PC5 & 0.0 & 0.0 & 0.0 & 0.1 & 99.7 & 0.9 & -0.3 & 0.1 & -0.8 & -0.6 \\
  PC6 & -0.0 & 0.0 & -0.1 & -0.2 & -0.8 & 99.6 & 0.5 & -0.5 & -0.2 & -0.4 \\
  PC7 & -0.0 & 0.0 & -0.1 & 0.0 & 0.5 & -0.4 & 98.9 & 3.1 & 0.7 & 1.6 \\
  PC8 & 0.0 & 0.0 & -0.2 & -0.0 & -0.2 & 0.5 & -3.2 & 98.4 & -4.5 & -1.5 \\
  PC9 & -0.0 & -0.0 & -0.0 & 0.0 & 0.6 & 0.1 & -0.7 & 4.6 & 96.9 & -10.7 \\
  PC10 & -0.0 & -0.0 & 0.1 & -0.1 & 0.3 & 0.1 & -1.2 & 1.5 & 8.6 & 92.7 \\
   \hline
\end{tabular}
\caption{Correlation between scores of PCA for the Celiac dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
\label{tab:pc-popres}
\end{table}

>>>>

\abstract{\textbf{Motivation:} Genome-wide genetic studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge for the genetic community, severely slowing down genetic analyses. This increase in data size has led to the emergence of a range of specialized software for every part of the analysis pipeline. Yet, it is often difficult to choose which piece of software to use and how to combine all these software. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of the RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of our packages by analyzing a simulated Genome-Wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} https://privefl.github.io/bigstatsr/ and https://privefl.github.io/bigsnpr/\\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK available either independently or incorporated within large framework (\cite{Danecek2011,Li2011,Purcell2007}). Similarly, genome-wide analysis quality control software have been developed such as PLINK and Bioconductor GWASTools (\cite{Gogarten2012}). Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA (\cite{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}). Then, GWAS analysis itself depends on the genotype format, e.g. ProbABEL or SNPTEST for dosage data (\cite{Aulchenko2010,Marchini2010}.  Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice (\cite{Vilhjalmsson2015,Euesden2015}). As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

@@

The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory. They use what we could call an hybrid solution: memory-mapping. Memory-mapping is used to access data, possibly stored on disk, as if it were in memory. This solution is made available within R through an object called ``big.matrix'', available in R package bigmemory (\cite{Kane2013}).

We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ (\cite{Nielsen2008}). With the two packages we developed, we made this solution available in R and in C++ via package Rcpp (\cite{Eddelbuettel2011}). The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R (\cite{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra (\cite{Zeng2017,RSpectra2016}).

Usually, memory-mapping provides seamless and faster access than standard read/write operations. When some element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. All this is managed by the Operating System so that it is seamless and efficient. It means that if you use chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on. Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.

@@

Fast read/write operations from/to bed/bim/fam PLINK files are available. One should use PLINK to convert data to this format. In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}). PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.

@@

Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}). The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R (\cite{Browning2009}). The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This algorithm doesn't use phasing and is very fast. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions (\cite{Chen2016}). XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices. For each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets. The training set is used to build a model for predicting missing data. The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of individuals that have been wrongly imputed for that particular SNP.

@@

The function based on randomized projections runs in linear time in all dimensions (\cite{Lehoucq1996}). FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes the multiplication based on the ``big.matrix'' format, which enables parallel computations.

Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed. For instance, thinning is mandatory when computing PCs of a genotype matrix (\cite{Abdellaoui2013}). There are different options to thin SNPs based on Linkage Disequilibrium. The first option is known as pruning, which is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning.

A variant of pruning is clumping. Clumping is useful if a statistic is available to sort the SNPs by importance (e.g. association with a phenotype) and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest. Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.

The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions (\cite{Price2008}). Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in these regions while computing PCs\footnote{https://goo.gl/8TngVE}. However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt (\cite{Luu2017}). Finally, the algorithm keeps only consecutive outlier SNPs which is considered as the sign of long-range LD regions by the algorithm. Indeed, a long-range LD region would cause SNPs in this region to have strong weights (loadings) in the PCA and we can differentiate these from true outliers because they are consecutive. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

For association purposes, statistical tests based on linear and logistic regressions are available. Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome (\cite{Zheng2012}).

$\forall j \in \{1, \dots, m\}$, 
\itemize{
\item for the linear regression: $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $m$ is the number of SNPs, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender).
\item for the logistic regression: $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $\hat{p} = \mathbb{P}(Y = 1)$.
}\\
The hypothesis that is tested is $\beta^{(j)} = 0$ against the alternative $\beta^{(j)} \neq 0$.

@@

First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations. Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients (\cite{Dudbridge2013,Chatterjee2013,Golan2014}).

Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors. The currently available models are linear and logistic regressions and Support Vector Machine (SVM). These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models (\cite{Tibshirani1996,Zou2005,Friedman2010}). Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors (\cite{Tibshirani2012}). The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso (\cite{Zeng2017}). These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).

@@

In this paper, we use two real datasets: the celiac disease cohort and the POPRES datasets (\cite{Dubois2010,Nelson2008}). The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.

For computation timing purposes only, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same structure of Linkage Disequilibrium and Population Structure as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset.

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compared the performance of the packages against standard software. Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: Celiac, a real case-control cohort, POPRES, a general population cohort and simulated datasets using real genotypes from the Celiac cohort. We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores. For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort. The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 (\cite{Purcell2007}). 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that we imputed with our new method (see section~\ref{sec:impute}). We note that if we used a standard R matrix to store the genotypes, it would take 32GB of memory. On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB.

@@

\subsection{Imputation of missing values for genotyped SNPs}\label{sec:impute}

@@

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Tue Jun 13 11:25:56 2017
\begin{table}[ht]
\centering
\begin{tabular}{l|cccccccccc}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 \\
  \hline
PC1 & 100.0 & -0.1 & -0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
  PC2 & 0.1 & 100.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 \\
  PC3 & 0.1 & -0.0 & 99.9 & 0.2 & -0.0 & 0.1 & 0.1 & 0.1 & 0.0 & -0.1 \\
  PC4 & -0.0 & -0.0 & -0.3 & 99.9 & -0.1 & 0.1 & -0.1 & 0.0 & 0.1 & 0.1 \\
  PC5 & 0.0 & 0.0 & 0.0 & 0.1 & 99.7 & 0.9 & -0.3 & 0.1 & -0.8 & -0.6 \\
  PC6 & -0.0 & 0.0 & -0.1 & -0.2 & -0.8 & 99.6 & 0.5 & -0.5 & -0.2 & -0.4 \\
  PC7 & -0.0 & 0.0 & -0.1 & 0.0 & 0.5 & -0.4 & 98.9 & 3.1 & 0.7 & 1.6 \\
  PC8 & 0.0 & 0.0 & -0.2 & -0.0 & -0.2 & 0.5 & -3.2 & 98.4 & -4.5 & -1.5 \\
  PC9 & -0.0 & -0.0 & -0.0 & 0.0 & 0.6 & 0.1 & -0.7 & 4.6 & 96.9 & -10.7 \\
  PC10 & -0.0 & -0.0 & 0.1 & -0.1 & 0.3 & 0.1 & -1.2 & 1.5 & 8.6 & 92.7 \\
   \hline
\end{tabular}
\caption{Correlation between scores of PCA for the Celiac dataset  when automatically removing long-range LD regions and when removing them based on a predefined table.}
\label{tab:pc-celiac}
\end{table}
